---
title: "datasets for SLAWR REU data vis"
author: "rachel torres"
date: "2025-06-19"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Data sets 

- already tidy 
- time series 
- categories with continuous 

1. Mercury levels in fish 1998 v. 2023 

Paragraph about mercury pollution and health 

Columns: 

- mercury content ug/g/
- length (cm)
- year 

2. Water quality time series comparing 2 streams 

Paragraph about water quality and fish habitat 

Columns:

- year 
- stream temperature (C)
- dissolved oxygen 
- sample site 

3. Temperature in Minnesota 

paragraph about climate change 

Columns: 

- year 
- month 
- min (degF)
- max (degF)

4. something about fish species 

inaturalist - query was: 

quality_grade=research&
identifications=any&
iconic_taxa[]=Actinopterygii&
place_id=124854&
verifiable=true&
threatened=true&
introduced=false&
native=true&
popular=true


```{r code}

library(tidyverse)

inat = read.csv("GitHub/slawr_reu_ieds/slawr-reu/data/observations-587687.csv/observations-587687.csv")

unique(inat$common_name)

ggplot(inat, aes(y=common_name, fill=place_state_name)) + geom_bar()

inat_tab = inat %>% select(place_state_name, common_name) %>% 
  table() %>% data.frame() %>% 
  filter(place_state_name == "Minnesota" |
           place_state_name == "Wisconsin") %>% 
  rename(count = Freq)

ggplot(inat_tab, aes(x=place_state_name, y=count)) +
  geom_boxplot()

ggplot(inat_tab, aes(x=place_state_name, 
                     y=common_name,
                     fill=count)) +
  geom_raster()

ggplot(inat_tab, aes(y=common_name, 
                     x=count, 
                     fill=place_state_name)) +
  geom_col()

  


```



